---
title: "Notes2 - Reproducible Research"
author: "Linda Kukolich"
date: "January 15, 2015"
output: html_document
---
# Coding Standards in R

1. Use text files and a text editor
2. Indent your code
3. Limit the width of your code to 80 columns
4. Limit the length of functions

*Seriously? Nothing on "Do not repeat code" or "use variable names that help 
explain the code"?*

# Markdown

> "Markdown is a text-to-HTML conversion tool for web writers. It allows you to
write using an easy-to-read, easy-to-write plain text format, and then convert
it to structurally valid XHTML (or HTML)."

**John Gruber, creator of Markdown**

Aaron Schwartz is also a creator of Markdown

```
# Level 1 header
## Level 2
### Level 3
```
# Level 1 header
## Level 2
### Level 3

```
* *italics*
* **Bold**
```

* *italics*
* **Bold**

```
- also a list
- with stuff
    a. this is also an ordered list
    b. with lower-alpha as a type
```
- also a list
- with stuff
    a. this is also an ordered list
    b. with lower-alpha as a type

```
[linkText](http://www.google.com)
```

Links can also be something like footnotes, when you need to reference the
same place multiple times.

```
I spend so much time reading [R bloggers][rb] and [Simply Statistics][ss]

[rb]: http://www.r-bloggers.com/ "R bloggers"

[ss]: http://simplystatistics.org/ "Simply Statistics"

And here is a second refers to [R bloggers][rb]
```

I spend so much time reading [R bloggers][rb] and [Simply Statistics][ss]

[rb]: http://www.r-bloggers.com/ "R bloggers"

[ss]: http://simplystatistics.org/ "Simply Statistics"

And here is a second refers to [R bloggers][rb]

Further information on Markdown is found at [dairingfireball.net]

# R Markdown

Allows mixing in blocks of R code for documentation and to generate plots.

Rmd becomes md becomes html

# knitr

## Literate Statistical Programming

- original idea from Donald Knuth
- An article is a stream of text and code.
- Analysis code is divided into text and code chunks
- Presentation code formats results (table, plots)
- Article text explains what is going on
- Literate programs are woven to produce human-readable documents and tangled
to produce machine readable documents
- Literate programming is a general concept with needs
    - a documentation language
    - a programming language
- **Sweave** by Friedrich Leisch used LaTeX and R
- **knitr** supports a variety of documentation languages (plot R)
    - written by Yihui Xie (while at Iowa State)
    - Supports RMarkdown, LaTeX, and HTML
    - can export to PDF, HTML
    - RStudio support

## How do I make my work reproducible?

- Decide to do it from (ideally from the beginning)
- Keep track of things (use version control to save intermediate state)
- Use software whose operation can be coded
- Don't save output
- Save data in non-proprietary formats

## Pros and Cons

- Pros
    - Store text and code all in one place
    - Data, results are automatically updated to reflect external changes
    - Code is live, an automatic regression test while building a document
- Cons
    - Text and code all in one place
    - If there is a lot of code, this can be hard to read
    - Can substantially slow down process of documents
- Things knitr is good for
    - Manuals
    - Short/medium-length technical documents
    - Tutorials
    - reports (especially if generated periodically)
    - Data preprocessing documents/summaries
- Things knitr is NOT good for
    - Very long reserch articles
    - complex time-consuming computations
    - Documents that require precise formatting
    
```
library(knitr)
setwd(<working directory>)
knit2html("document.Rmd")
browseURL("document.html")
```

## Cool stuff
{r simulation, echo = FALSE, results="hide"} to supress not just the source
code, but also the prints. Not, however, the plots. The simulation part seems
to be a code block label, though I do not know what use it has in either
RMarkdown or in the final HTML. It is not in the HTML.

```{r simulation, echo = FALSE, results="hide"}
x <- rnorm(100)
mean(x)
hist(x)
```

Inline: The current time is `r date()`. My favorite random number is 
`r sample(1:100, 1)`

Tables:
```{r fitmodel}
library(datasets)
data(airquality)
fit <- lm(Ozone ~ Wind + Temp + Solar.R, data = airquality)
```

Here is a table of regression coefficients
```{r showtable, results="asis"}
library(xtable)
xt <- xtable(summary(fit))
print(xt, type = "html")
```

### Setting Global defaults

```
opts_chunk$set(echo = FALSE, results = "hide")
```

Settings

- results: asis, hide
- echo: TRUE, FALSE
- fig.height, fig.width: numeric
- cache - store results after first pass and then reload later
    - if the data or code or anything external changes, will need to rerun
    - dependencies are not checked
    - chucks with side effects may not be cacheable
    

# Peer Assessment 1

