---
title: "Biostatistics BootCamp II: Error on Slide: Lecture 4 - Slide 13"
author: "Linda Kukolich"
date: "February 4, 2015"
output: html_document
---
There seems to be an error on one of the slides in lecture 4. Specifically, the result of the calculation given on the slide:
```{r}
(.55 - .25)/sqrt(.4*.6*(2/20))
```

does not equal the result given: 1.61

## More detail

The Data, taken from Lecture 4, Slide 3
```{r}
df <- data.frame(SideEffects = c(11, 5, 16), None=c(9, 15, 14), total=c(20, 20, 40), row.names=c("A", "B", "Total"))
```

***Text of the slide in question***

## Example

- Test whether or not the proportion of side effects is the same for the two drugs
- $\hat{p}_A = .55, \hat{p}_B = 5/20 = .25, \hat{p} = 16/40 = .4$
- Test statistic

$$
\frac{.55 - .25}{\sqrt{.4 \times .6 \times (1/20 + 1/20)}} = 1.61
$$

- Fail to reject $H_0$ at .05 level (compare with 1.96)
- P-value $P(|Z| \ge 1.61) = .11$

***My Numbers Do Not Match Dr Caffo's***
His can be approximated with .25/sqrt(.024).

## My calculations
```{r}
# My calculations
n_a <- df["A", "total"]
phata <- df["A", "SideEffects"] / n_a
n_b <- df["B", "total"]
phatb <- df["B", "SideEffects"] / n_b
n <- df["Total", "total"]
phat <- df["Total", "SideEffects"] / n
ts <- (phata - phatb)/sqrt(phat * (1-phat) * (1/n_a + 1/n_b))
ts
```
```{r}
z975 <- qnorm(.975)
if (abs(ts) >= z975) {
    print(paste(ts, ">=", z975, "Reject null hypothesis"))
} else {
    print(paste(ts, "<", z975, "Fail to reject null hypothesis"))
}
pvalue <- 2*pnorm(ts, lower.tail=FALSE)
pvalue
```

## Possible calculation that gets Dr Caffo's result
```{r}
ts <- phatb/sqrt(phat * (1-phat) * (1/n_a + 1/n_b))
ts
z975 <- qnorm(.975)
if (abs(ts) >= z975) {
    print(paste(ts, ">=", z975, "Reject null hypothesis"))
} else {
    print(paste(ts, "<", z975, "Fail to reject null hypothesis"), digits=2)
}
pvalue <- 2*pnorm(ts, lower.tail=FALSE)
print(pvalue, digits=2)
```
