---
title: "R-Equations"
author: "Linda Kukolich"
date: "January 31, 2015"
output: html_document
---

# error types
- Type 1 error - take $H_a$ when $H_0$ was the truth, false alarm
- Type 2 error - take $H_0$ when $H_a$ was the truth, miss

# Z-test - high number of samples, so Central Limit Theorem applies (100 or more)

How many standard deviations from the population mean is our sample mean?

- Test statistic $TS = \frac{\bar{X} - \mu_0}{S/\sqrt{n}}$
- Reject the null hypothesis when
    - $H_1 : TS \le - Z_{1-\alpha}$
    - $H_2 : |TS| \ge Z_{1-\alpha/2}$
    - $H_3 : TS \ge Z_{1-\alpha}$

## One sided z-test

```{r}
mu <- 30
xbar <- 32
sd <- 10
n <- 100
ts <- (xbar - mu)/(sd/sqrt(n))
z95 <- qnorm(1-0.05)
if(ts >= z95) {
    print(paste(ts, ">=", z95))
    print("reject null hypothesis (accept Ha)")
} else if (ts <= -z95) {
    print(paste(ts, "<=", -z95))
    print("reject null hypothesis (accept Ha)")
} else {
    print(paste(ts, "<", z95))
    print("fail to reject null hypothesis")
}
```

## two sided z-test

```{r}
mu <- 30
xbar <- 32
sd <- 10
n <- 100
ts <- (xbar - mu)/(sd/sqrt(n))
z975 <- qnorm(1-0.025)
if(abs(ts) >= z975) {
    print(paste(abs(ts), ">=", z975))
    print("reject null hypothesis (accept Ha)")
} else {
    print(paste(ts, "<", z975))
    print("fail to reject null hypothesis")
}
```

## Z-test Confidence interval
```{r}
mu0 <- 30
xbar <- 32
sd <- 10
n <- 100
ts <- (xbar - mu0)/(sd/sqrt(n))
# we are 95% sure that mu0 lies within this confidence interval
conf <- xbar + c(-1, 1) * qnorm(1-0.05) * sd/sqrt(n)
if ((mu0 >= conf[1]) && (mu0 <= conf[2])) {
    print(paste("mu0 (", mu0, ") lies within this confidence interval: (", conf[1], ",", conf[2], ")"))
} else {
    print(paste("We find that mu0 (", mu0, ") lies outside this 95% confidence interval: (", conf[1], ",", conf[2], ")"))
}
```

## Z-test power

```{r}
mu0 <- 30
xbar <- 32
n <- 100
sd <- 10
z95 <- qnorm(1-0.05)
ts <- (xbar-mu0)/(sd/sqrt(n))
power <- 1- pnorm(z95-ts)
power
```

## Sample size calculation (power 80%)
*works for mua or larger, mua or smaller is similar*
```{r}
mua = 32
mu0 = 30
sigma = 10
z95 = qnorm(1-0.05)
z20 = qnorm(1-.80)
((z95 - z20) * sigma / (mua - mu))^2
```

## two sided sample size calculation
```{r}
mua = 32
mu0 = 30
sigma = 10
z975 = qnorm(1-0.025)
z20 = qnorm(1-.80)
((z975 - z20) * sigma / (mua - mu))^2
```

# T-test - low number of samples

## One sided t test

```{r}
mu <- 30
xbar <- 32
sd <- 10
n <- 16 # Too small for Z-test
ts <- (xbar - mu)/(sd/sqrt(n))
t95 <- qt(1-0.05, n-1)
if(ts >= t95) {
    print(paste(ts, ">=", t95))
    print("reject null hypothesis (accept Ha)")
} else if (ts <= -t95) {
    print(paste(ts, "<=", -t95))
    print("reject null hypothesis (accept Ha)")
} else {
    print(paste(ts, "<", t95))
    print("fail to reject null hypothesis")
}
```

## Two sided t test
```{r}
mu <- 30
xbar <- 32
sd <- 10
n <- 16 # Too small for Z-test
ts <- (xbar - mu)/(sd/sqrt(n))
t975 <- qt(1-0.025, n-1)
if(abs(ts) >= t975) {
    print(paste(abs(ts), ">=", t975))
    print("reject null hypothesis")
} else {
    print(paste(ts, "<", t975))
    print("fail to reject null hypothesis (accept Ha)")
}
```

## Student's T confidence interval
A sample of 9 men yielded a sample average brain volume of 1,100cc and a standard deviation of 30cc. For what values of μ0 would a test of H0:μ=μ0 fail to reject the null hypothesis in a two sided 5% Students t-test?

```{r}
mu <- 1100
df <- 8
sd <- 30
t975 <- qt(1-.025, n-1)
mu + c(-1, 1) * t975 * sd/sqrt(n)
```

##  T-test Power

Probability of seeing evidence as or more extreme than actually obtained is

```{r}
mu0 <- 30
mua <- 32
sd <- 4
n <- 16 # Too small for Z-test
power.t.test(n = n, delta = (mua - mu0)/sigma, type = "one.sample", alt = "one.sided")
power.t.test(n = n, delta = (mua - mu0), sd = sd, type = "one.sample", alt = "one.sided")
```

# Two sample tests - two independent groups

### With raw data
A pharmaceutical company is interested in testing a potential blood pressure lowering medication. Their first examination considers only subjects that received the medication at baseline then two weeks later. The data are as follows (SBP in mmHg)

| Subject |   Baseline |    Week 2 |
|---------|------------|-----------|
| 1	| 140	| 132 | 
| 2	| 138 | 	135 | 
| 3	| 150 | 	151 | 
| 4	| 148	| 146 | 
| 5	| 135	| 130 | 

```{r}
bl <- c(140, 138, 150, 148, 135)
fu <- c(132, 135, 151, 146, 130)
t.test(fu, bl, alternative = "two.sided", paired=TRUE)
```

```{r}
diff <- fu - bl
n <- length(fu)
m <- mean(diff)
s <- sd(diff)
ts <- sqrt(n) * m / s
t975 <- qt(.975, n - 1)
if (abs(ts) >= t975) {
    print(paste(abs(ts), ">=", t975))
    print("reject null hypothesis (accept means are not equal)")
} else {
    print(paste(abs(ts), "<", t975))
    print("fail to reject null hypothesis (means are the same)")
}
# confidence interval, matched data
conf <- m + c(-1, 1) * t975 * s * sqrt(1/n)
if ((conf[1] < 0) && (conf[2] > 0)) {
    print(paste("0 lies within this confidence interval: (", conf[1], ",", conf[2], "), so means are the same"))
} else {
    print(paste("0 lies outside this 95% confidence interval: (", conf[1], ",", conf[2], "), so means are different"))
}
# power - matched data
2*pt(abs(ts), n - 1, lower.tail = FALSE)
```

```{r}
# Shared variance
# H0: mu1 = mu2, versus Ha: mu1 <> mu2
xbar <- mean(fu)
nx <- length(fu)
sx <- sd(fu)
c(xbar, sx)

ybar <- mean(bl)
ny <- length(bl)
sy <- sd(bl)
c(ybar, sy)

sp <- sqrt(((nx-1)*sx^2 + (ny-1)*sy^2)/(nx + ny - 2))
sp
# pooled (shared) variance
ts <- (xbar - ybar)/(sp * sqrt(1/nx + 1/nx))
df <- nx + ny - 2
t975 <- qt(.975, df)
if (abs(ts) >= t975) {
    print(paste(abs(ts), ">=", t975))
    print("reject null hypothesis (accept means are not equal)")
} else {
    print(paste(abs(ts), "<", t975))
    print("fail to reject null hypothesis (means are the same)")
}
```
```{r}
# confidence interval, pooled variance
conf <- (xbar - ybar) + c(-1, 1) * t975 * sp * sqrt(1/nx + 1/ny)
if ((conf[1] < 0) && (conf[2] > 0)) {
    print(paste("0 lies within this confidence interval: (", conf[1], ",", conf[2], "), so means are the same"))
} else {
    print(paste("0 lies outside this 95% confidence interval: (", conf[1], ",", conf[2], "), so means are different"))
}
# Power, pooled variance
2*pt(abs(ts), df, lower.tail = FALSE)
```
```{r}
t.test(fu, bl, alternative = "two.sided", paired=FALSE)
```
```{r}
# not pooled (shared) variance
ts <- (xbar - ybar)/sqrt(sx^2/nx + sy^2/ny)
ts
df <- (sx^2/nx + sy^2/ny)^2/((sx^2/nx)^2/(nx-1) + (sy^2/ny)^2/(ny-1))
df
t975 <- qt(.975, df)
if (abs(ts) >= t975) {
    print(paste(abs(ts), ">=", t975))
    print("reject null hypothesis (accept means are not equal")
} else {
    print(paste(abs(ts), "<", t975))
    print("fail to reject null hypothesis (means are the same)")
}
```
```{r}
# Confidence interval, not pooled variance
conf <- xbar - ybar + c(-1, 1) * t975 * sqrt(sx^2/nx + sy^2/ny)
if ((conf[1] < 0) && (conf[2] > 0)) {
    print(paste("0 lies within this confidence interval: (", conf[1], ",", conf[2], "), so means are the same"))
} else {
    print(paste("0 lies outside this 95% confidence interval: (", conf[1], ",", conf[2], "), so means are different"))
}
# power, not pooled variance
2*pt(abs(ts), df, lower.tail = FALSE)
```
